---
title: "Causal ML for Treatment Outcomes: Smoking, BMI, and Diabetes"
subtitle: "Illustrating Feuerriegel et al. (2024) *Nature Medicine* with the G-Formula and Causal Mediation"
author: "Arda"
date: today
date-format: long
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    highlight-style: github
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    embed-resources: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

## Motivation

@feuerriegel2024 argue that standard ML pipelines fail when estimating
treatment effects because they ignore the **causal structure** linking
treatment, mediators, and outcomes. Their canonical example:

> *"Consider a patient with a high body mass index whose doctor
> recommends quitting smoking, and for whom the diabetes risk should be
> predicted. Literature from traditional ML would suggest using both the
> body mass index and smoking behavior to predict the diabetes risk
> under a smoking versus no-smoking scenario; however, this approach
> would ignore that stopping smoking would also change a patient's body
> mass index."*

This document implements a principled causal analysis of exactly that
example using three approaches:

| Approach                 | Description                                                    | Validity                           |
|--------------------------|----------------------------------------------------------------|------------------------------------|
| **(A) Naïve ML**         | Logistic regression on post-treatment BMI                      | ❌ Biased (conditions on mediator) |
| **(B) Causal mediation** | Decomposes total effect into direct + BMI-mediated [@imai2010] | ✅ Correct decomposition           |
| **(C) G-formula / SCM**  | Intervenes `do(quit=1)`, propagates through BMI                | ✅ Correct counterfactual ATE      |

### Empirical basis

The synthetic cohort is calibrated to three published sources:

-   @hu2018 (NHS / NHS-II / HPFS cohorts, *NEJM*): post-cessation weight
    gain explained \~68% of the transient type 2 diabetes (T2D) spike;
    HR ≈ 1.91 for quitters with ≥10 kg gain.
-   **Kangbuk Samsung Health Study** [-@kangbuk2024]: mean BMI 25.68 ±
    3.53 (T2D) vs 23.18 ± 3.28 (non-T2D); current smokers HR ≈ 1.29.
-   @bush2016: post-cessation BMI increase ≈ 0.9 kg/m² within 3 years;
    magnitude varies with baseline BMI.

------------------------------------------------------------------------

## Packages

```{r packages}
pkgs <- c("ggplot2", "mediation")
for (p in pkgs) if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
library(ggplot2)
library(mediation)
```

------------------------------------------------------------------------

## Causal DAG

The assumed structural equation model:

$$
\begin{aligned}
\text{BMI}_0 &= f(\text{Age, Sex, FamHx, Alcohol, Activity}) + \varepsilon_1 \\
\text{Quit}  &= \text{Bernoulli}\!\left(\sigma(\text{Age, Sex, BMI}_0, \text{Activity})\right) \\
\Delta\text{BMI} &= 0.90\,\text{Quit} + 0.05\,(\text{BMI}_0-23)\cdot\text{Quit} + \cdots + \varepsilon_2 \\
\text{T2D}   &= \text{Bernoulli}\!\left(\sigma(1.5\,\Delta\text{BMI} - 0.30\,\text{Quit} + \cdots)\right)
\end{aligned}
$$

The key path ignored by naïve ML is **Quit → ΔBMI → T2D**.

```{r dag-plot, fig.cap="Causal DAG. The mediated path (green) through ΔBMI is what naïve ML ignores when it conditions on post-treatment BMI.", fig.height=4.5}
dag_nodes <- data.frame(
  node = c("Quit\nSmoking", "ΔBMI", "Diabetes", "Baseline\nBMI",
           "Age/Sex/\nFam Hx", "Alcohol/\nActivity"),
  x    = c(3, 5, 7, 3, 1, 1),
  y    = c(3, 3, 3, 5, 5, 1)
)

dag_edges <- data.frame(
  from = c("Quit\nSmoking","ΔBMI","Baseline\nBMI","Baseline\nBMI",
           "Age/Sex/\nFam Hx","Age/Sex/\nFam Hx","Alcohol/\nActivity",
           "Alcohol/\nActivity","Quit\nSmoking"),
  to   = c("ΔBMI","Diabetes","ΔBMI","Diabetes",
           "Quit\nSmoking","Diabetes","ΔBMI","Diabetes","Diabetes"),
  type = c("mediated","mediated","conf","conf",
           "conf","conf","conf","conf","direct")
)

dag_coords <- setNames(dag_nodes[, c("x","y")], c("x","y"))
rownames(dag_coords) <- dag_nodes$node

edges_plot <- do.call(rbind, lapply(seq_len(nrow(dag_edges)), function(i) {
  f <- dag_edges$from[i]; t <- dag_edges$to[i]
  data.frame(x    = dag_coords[f,"x"], y    = dag_coords[f,"y"],
             xend = dag_coords[t,"x"], yend = dag_coords[t,"y"],
             type = dag_edges$type[i])
}))

ggplot() +
  geom_segment(
    data  = edges_plot,
    aes(x = x, y = y, xend = xend, yend = yend, colour = type),
    arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
    linewidth = 0.9, alpha = 0.85
  ) +
  geom_label(
    data  = dag_nodes,
    aes(x = x, y = y, label = node),
    size  = 3.2, fill = "white", colour = "#1a1a1a",
    label.size = 0.4, label.padding = unit(0.3, "lines")
  ) +
  scale_colour_manual(
    values = c("mediated" = "#1D9E75", "direct" = "#3B8BD4", "conf" = "#888780"),
    labels = c("mediated" = "Mediated path (Quit → ΔBMI → T2D)",
               "direct"   = "Direct effect (Quit → T2D)",
               "conf"     = "Confounder / covariate")
  ) +
  annotate("text", x = 4, y = 3.32,
           label = "Mediator (BMI gain)", size = 3, colour = "#085041") +
  annotate("text", x = 5, y = 2.65,
           label = "Ignored by\nnaïve ML", size = 2.8,
           colour = "#993C1D", fontface = "italic") +
  labs(title   = "Causal DAG: Smoking cessation → ΔBMI → Diabetes",
       subtitle = "Hu et al. (2018) NEJM; Kangbuk Samsung cohort (2024)",
       colour = NULL, x = NULL, y = NULL) +
  theme_void(base_size = 11) +
  theme(plot.title      = element_text(face = "bold"),
        legend.position = "bottom",
        legend.text     = element_text(size = 9))
```

------------------------------------------------------------------------

## Synthetic cohort

```{r data-gen}
set.seed(2024)
n <- 2000

# Baseline covariates
age      <- round(rnorm(n, 50, 10))
sex      <- rbinom(n, 1, 0.45)          # 1 = male
fam_hx   <- rbinom(n, 1, 0.20)         # family history of T2D
alcohol  <- pmax(0, rnorm(n, 8, 5))    # g/day
activity <- rbinom(n, 1, 0.35)         # physically active

# Baseline BMI — smokers slightly leaner (MR evidence, Atkins 2021)
bmi0 <- 23.5 +
  0.05 * (age - 50) + 1.20 * sex + 0.50 * fam_hx +
  0.04 * alcohol   - 0.80 * activity + rnorm(n, sd = 3.0)
bmi0 <- pmax(16, bmi0)

# Treatment: quit smoking
lp_quit <- -1.2 + 0.015*(age-50) - 0.20*sex + 0.02*(bmi0-23) + 0.30*activity
quit    <- rbinom(n, 1, plogis(lp_quit))

# Mediator: post-cessation BMI change
# Calibrated to Bush (2016): mean gain ≈ +0.9 kg/m² for quitters
delta_bmi <- 0.90 * quit +
             0.05 * (bmi0 - 23) * quit +   # more gain at higher baseline BMI
             0.30 * (1 - activity) * quit - # sedentary quitters gain more
             0.15 * activity + rnorm(n, sd = 1.2)

bmi_post <- bmi0 + delta_bmi

# Outcome: incident T2D (~5-year risk)
# Calibrated to HR ≈ 1.29 (smokers vs never, Kangbuk Samsung 2024)
# and BMI OR ≈ 1.10 per unit (Hu 2018)
lp_dm <- -5.50 +
  0.035 * (age - 50) + 0.50 * sex + 0.80 * fam_hx +
  0.20  * (bmi_post - 23) + 0.025 * alcohol - 0.40 * activity +
 -0.30  * quit + rnorm(n, sd = 0.3)

diabetes <- rbinom(n, 1, plogis(lp_dm))

cohort <- data.frame(age, sex, fam_hx, alcohol, activity,
                     bmi0, quit, delta_bmi, bmi_post, lp_dm, diabetes)
```

### Cohort summary

```{r cohort-summary}
summary_tbl <- data.frame(
  Statistic = c(
    "Sample size",
    "Quitters, n (%)",
    "T2D cases, n (%)",
    "Mean baseline BMI (SD)",
    "Mean post-treatment BMI (SD)",
    "Mean ΔBMI — quitters",
    "Mean ΔBMI — non-quitters",
    "T2D rate — quitters",
    "T2D rate — non-quitters"
  ),
  Value = c(
    n,
    sprintf("%d (%.1f%%)", sum(quit), 100*mean(quit)),
    sprintf("%d (%.1f%%)", sum(diabetes), 100*mean(diabetes)),
    sprintf("%.2f (%.2f)", mean(bmi0), sd(bmi0)),
    sprintf("%.2f (%.2f)", mean(bmi_post), sd(bmi_post)),
    sprintf("+%.2f kg/m²", mean(delta_bmi[quit==1])),
    sprintf("%.2f kg/m²",  mean(delta_bmi[quit==0])),
    sprintf("%.1f%%", 100*mean(diabetes[quit==1])),
    sprintf("%.1f%%", 100*mean(diabetes[quit==0]))
  )
)
knitr::kable(summary_tbl, col.names = c("Statistic", "Value"),
             caption = "Cohort descriptive statistics")
```

------------------------------------------------------------------------

## Analysis A --- Naïve ML (incorrect)

Logistic regression predicting T2D from `(quit, bmi_post, covariates)`.
This **conditions on `bmi_post`**, a post-treatment variable that lies
on the causal path from `quit` to `diabetes`. Doing so blocks the
mediated pathway and underestimates the causal effect of quitting.

```{r naive-ml}
naive_fit <- glm(
  diabetes ~ quit + bmi_post + age + sex + fam_hx + alcohol + activity,
  data = cohort, family = binomial
)
naive_coef <- coef(summary(naive_fit))

knitr::kable(
  data.frame(
    Term    = rownames(naive_coef),
    Estimate = round(naive_coef[,"Estimate"], 4),
    OR       = round(exp(naive_coef[,"Estimate"]), 3),
    `p-value` = round(naive_coef[,"Pr(>|z|)"], 4)
  ),
  caption = "Naïve ML: logistic regression coefficients (biased)"
)
```

::: callout-warning
## Why this is wrong

`bmi_post` is a **mediator**, not a baseline covariate. Including it in
the outcome model as if it were fixed --- independently of `quit` ---
blocks the `quit → ΔBMI → T2D` path. The resulting OR for `quit`
understates the true causal effect, because part of quitting's impact
(operating through weight gain) is absorbed into the `bmi_post`
coefficient.
:::

------------------------------------------------------------------------

## Analysis B --- Causal mediation analysis

Using the `mediation` package [@imai2010] to decompose the total effect
of quitting into:

$$\text{TE} = \underbrace{\text{ACME}}_{\text{via ΔBMI}} + \underbrace{\text{ADE}}_{\text{direct}}$$

```{r mediation, cache=TRUE}
# Mediator model: ΔBMI ~ quit + confounders
med_model <- lm(
  delta_bmi ~ quit + bmi0 + age + sex + fam_hx + alcohol + activity,
  data = cohort
)

# Outcome model: T2D ~ quit + ΔBMI + confounders
out_model <- glm(
  diabetes ~ quit + delta_bmi + bmi0 + age + sex + fam_hx + alcohol + activity,
  data = cohort, family = binomial
)

med_out <- mediate(
  med_model, out_model,
  treat    = "quit",
  mediator = "delta_bmi",
  robustSE = TRUE,
  sims     = 500
)
```

```{r mediation-results}
med_tbl <- data.frame(
  Component = c("ACME (BMI-mediated)", "ADE (direct effect)", "Total effect",
                "Proportion mediated"),
  Estimate  = c(med_out$d0,      med_out$z0,      med_out$tau.coef, med_out$n0),
  CI_low    = c(med_out$d0.ci[1],med_out$z0.ci[1],med_out$tau.ci[1], NA),
  CI_high   = c(med_out$d0.ci[2],med_out$z0.ci[2],med_out$tau.ci[2], NA)
)
knitr::kable(med_tbl, digits = 4,
             caption = "Causal mediation decomposition (Imai et al. 2010)")
```

```{r mediation-plot, fig.cap="Decomposition of the total effect of quitting smoking on T2D risk into the BMI-mediated component (ACME) and the direct effect (ADE). Error bars are 95% bootstrap CIs."}
med_bar <- data.frame(
  component = factor(c("ACME\n(BMI-mediated)", "ADE\n(direct effect)"),
                     levels = c("ACME\n(BMI-mediated)", "ADE\n(direct effect)")),
  estimate  = c(med_out$d0, med_out$z0),
  lo        = c(med_out$d0.ci[1], med_out$z0.ci[1]),
  hi        = c(med_out$d0.ci[2], med_out$z0.ci[2])
)

ggplot(med_bar, aes(x = component, y = estimate, fill = component)) +
  geom_col(width = 0.45, alpha = 0.85) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.12, linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
  scale_fill_manual(values = c("#1D9E75", "#3B8BD4")) +
  labs(
    title    = "Causal mediation: total effect decomposition",
    subtitle = sprintf("Proportion mediated via BMI: %.1f%%  |  Total ATE: %.4f",
                       med_out$n0 * 100, med_out$tau.coef),
    x = NULL, y = "Effect on P(T2D = 1)", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "none")
```

------------------------------------------------------------------------

## Analysis C --- G-formula (parametric SCM)

The fully causal approach. For each individual we:

1.  Fit $E[\Delta\text{BMI} \mid \text{do(quit}=t\text{)}, \mathbf{X}]$
    --- the mediator model.
2.  Predict $\widehat{\Delta\text{BMI}}$ under the intervention
    $t \in \{0,1\}$.
3.  Plug $\widehat{\Delta\text{BMI}}$ into the outcome model to predict
    T2D risk.
4.  Average over the covariate distribution $\rightarrow$ ATE.

This implements $E[Y(\text{do}(A=1))]$ rather than $E[Y \mid A=1]$.

```{r gformula, cache=TRUE}
med_gform <- lm(
  delta_bmi ~ quit + bmi0 + age + sex + fam_hx + alcohol + activity,
  data = cohort
)
out_gform <- glm(
  diabetes ~ quit + delta_bmi + bmi0 + age + sex + fam_hx + alcohol + activity,
  data = cohort, family = binomial
)

gformula_ate <- function(df, treat_val) {
  df_int           <- df
  df_int$quit      <- treat_val
  df_int$delta_bmi <- predict(med_gform, newdata = df_int)
  mean(predict(out_gform, newdata = df_int, type = "response"))
}

p_dm_quit  <- gformula_ate(cohort, 1)
p_dm_smoke <- gformula_ate(cohort, 0)
ate_gform  <- p_dm_quit - p_dm_smoke

# Bootstrap 95% CI
set.seed(42)
boot_ate <- replicate(500, {
  idx <- sample(nrow(cohort), replace = TRUE)
  b   <- cohort[idx, ]
  m   <- lm(delta_bmi ~ quit + bmi0 + age + sex + fam_hx + alcohol + activity, data = b)
  o   <- glm(diabetes  ~ quit + delta_bmi + bmi0 + age + sex + fam_hx + alcohol + activity,
             data = b, family = binomial)
  f   <- function(tv) {
    d <- b; d$quit <- tv; d$delta_bmi <- predict(m, newdata = d)
    mean(predict(o, newdata = d, type = "response"))
  }
  f(1) - f(0)
})
ci_gform <- quantile(boot_ate, c(0.025, 0.975))
```

```{r gformula-results}
knitr::kable(
  data.frame(
    Quantity  = c("P(T2D = 1 | do(quit=1))",
                  "P(T2D = 1 | do(quit=0))",
                  "ATE  [do(quit=1) − do(quit=0)]",
                  "95% bootstrap CI (lower)",
                  "95% bootstrap CI (upper)"),
    Value = round(c(p_dm_quit, p_dm_smoke, ate_gform,
                    ci_gform[1], ci_gform[2]), 4)
  ),
  caption = "G-formula results: causal ATE of smoking cessation on T2D"
)
```

------------------------------------------------------------------------

## Naïve vs causal ATE comparison

```{r ate-comparison}
naive_pred_quit  <- mean(predict(naive_fit,
  newdata = transform(cohort, quit = 1), type = "response"))
naive_pred_smoke <- mean(predict(naive_fit,
  newdata = transform(cohort, quit = 0), type = "response"))
ate_naive <- naive_pred_quit - naive_pred_smoke

knitr::kable(
  data.frame(
    Method   = c("Naïve ML (biased — conditions on mediator)",
                 "G-formula (causal — propagates quit → ΔBMI → T2D)",
                 "Absolute bias"),
    ATE = round(c(ate_naive, ate_gform, abs(ate_naive - ate_gform)), 4)
  ),
  caption = "Naïve vs causal ATE: the bias introduced by ignoring the smoking → BMI pathway"
)
```

```{r scatter-plot, fig.cap="Individual-level naïve vs causal predicted T2D risk for quitters. Points above the 45° line (dashed) are patients for whom causal reasoning gives a *higher* risk estimate than naïve ML — correctly accounting for the BMI gain caused by quitting. Red colour indicates higher baseline BMI."}
cohort$risk_naive <- predict(naive_fit, type = "response")

tmp_q1           <- cohort; tmp_q1$quit <- 1
tmp_q1$delta_bmi <- predict(med_gform, newdata = tmp_q1)
cohort$risk_causal_quit <- predict(out_gform, newdata = tmp_q1, type = "response")

quitters <- cohort[cohort$quit == 1, ]

ggplot(quitters, aes(x = risk_naive, y = risk_causal_quit, colour = bmi0)) +
  geom_point(alpha = 0.4, size = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "#333") +
  scale_colour_gradient(low = "#3B8BD4", high = "#D85A30", name = "Baseline BMI") +
  labs(
    title    = "Naïve ML vs causal (g-formula) T2D risk — quitters only",
    subtitle = "Points above diagonal: causal model gives higher risk\n(propagates quit → BMI gain → diabetes)",
    x = "Naïve ML predicted risk  [conditions on observed post-BMI]",
    y = "Causal predicted risk  [do(quit=1), propagates ΔBMI]"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

------------------------------------------------------------------------

## Heterogeneous treatment effects (CATE)

@feuerriegel2024 emphasise **conditional** ATE (CATE). Patients with
high baseline BMI face greater weight gain after quitting, and thus a
larger short-term T2D risk increase --- even though quitting is
beneficial long-run [@hu2018].

```{r cate}
cohort$bmi_grp <- cut(
  bmi0,
  breaks = c(-Inf, 22, 25, 28, Inf),
  labels = c("< 22", "22–25", "25–28", "≥ 28")
)

cate_df <- do.call(rbind, lapply(levels(cohort$bmi_grp), function(g) {
  sub <- cohort[cohort$bmi_grp == g, ]
  if (nrow(sub) < 30) return(NULL)
  m  <- lm(delta_bmi ~ quit + bmi0 + age + sex + fam_hx + alcohol + activity, data = sub)
  o  <- glm(diabetes  ~ quit + delta_bmi + bmi0 + age + sex + fam_hx + alcohol + activity,
            data = sub, family = binomial)
  f  <- function(tv) {
    d <- sub; d$quit <- tv; d$delta_bmi <- predict(m, newdata = d)
    mean(predict(o, newdata = d, type = "response"))
  }
  p1 <- f(1); p0 <- f(0)
  data.frame(bmi_grp = g, n = nrow(sub),
             p_smoke = p0, p_quit = p1, cate = p1 - p0)
}))

knitr::kable(cate_df, digits = 4,
             caption = "CATE by baseline BMI group (g-formula)")
```

```{r cate-plot, fig.cap="CATE by BMI group. Red bars indicate a net increase in short-term T2D risk from quitting, concentrated in patients with higher baseline BMI."}
cate_df$bmi_grp <- factor(cate_df$bmi_grp, levels = levels(cohort$bmi_grp))

ggplot(cate_df, aes(x = bmi_grp, y = cate, fill = cate < 0)) +
  geom_col(width = 0.55, alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(
    values = c("FALSE" = "#D85A30", "TRUE" = "#1D9E75"),
    labels = c("FALSE" = "Quitting increases T2D risk (short-term)",
               "TRUE"  = "Quitting reduces T2D risk")
  ) +
  labs(
    title    = "CATE by baseline BMI group (g-formula)",
    subtitle = "High-BMI patients face greater short-term T2D risk from cessation\nvia larger post-cessation weight gain",
    x    = "Baseline BMI group (kg/m²)",
    y    = "CATE: P(T2D | do(quit=1)) − P(T2D | do(quit=0))",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

------------------------------------------------------------------------

## Summary

```{r summary-table}
knitr::kable(
  data.frame(
    Finding = c(
      "Naïve ML ATE (biased)",
      "G-formula ATE (causal)",
      "Absolute bias",
      "Proportion of total effect mediated via ΔBMI",
      "Direct protective effect of quitting (ADE)"
    ),
    Value = c(
      sprintf("%.4f", ate_naive),
      sprintf("%.4f  [95%% CI: %.4f, %.4f]", ate_gform, ci_gform[1], ci_gform[2]),
      sprintf("%.4f", abs(ate_naive - ate_gform)),
      sprintf("%.1f%%", med_out$n0 * 100),
      sprintf("%.4f  [95%% CI: %.4f, %.4f]", med_out$z0,
              med_out$z0.ci[1], med_out$z0.ci[2])
    )
  ),
  caption = "Key results"
)
```

::: callout-note
## Key takeaway

Naïve ML implicitly assumes BMI is unaffected by the treatment decision.
The g-formula correctly propagates `do(quit=1)` through the mediator
ΔBMI, giving a different --- and causally valid --- population-average
risk prediction. The CATE analysis further reveals that the short-term
T2D risk from quitting is **concentrated in higher-BMI patients**,
exactly the kind of clinically actionable heterogeneity that causal ML
can surface but associational ML misses.
:::

------------------------------------------------------------------------

## References

::: {#refs}
:::
